---
title: "RNASeq analysis"
output: html_document
date: "2025-02-04"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## starting info

setwd("/Users/mckenzir/Desktop/Others/RNAseq")
if (!require("BiocManager", quietly = TRUE))
  install.packages("BiocManager")

BiocManager::install("DESeq2")
library(DESeq2)
library(tidyverse)

#https://github.com/kpatel427/YouTubeTutorials/blob/main/runDESeq2.R

read.csv('counts_matrix.csv') -> counts_matrix
read.csv('filtered_counts_matrix.csv') -> filtered_counts_matrix
read.csv('sample_info.csv') -> sample_info

#Comes with X in front of the names so chose to load the R data files prepared by Andrew instead
load("~/Desktop/Others/RNAseq/counts_matrix.RData")
load("~/Desktop/Others/RNAseq/filtered_counts_matrix.RData")

#The row names were 1-24 so they needed to be changed to the actual names of the samples

rownames(sample_info) <- c('1b_1h_IL12','1h_1h_IL12','1b_1h_IL18','1h_6h_null','1b_6h_IL18','1e_6h_IL18',
                           '1e_1h_IL18','1h_6h_IL18','1b_6h_IL12','1e_6h_null','1h_1h_IL18','1b_6h_null',
                           '1e_1h_IL12','1h_6h_IL12','1e_1h_null','1b_1h_null','1h_1h_aCD3','1e_6h_aCD3',
                           '1h_6h_aCD3','1h_1h_null','1e_1h_aCD3','1b_1h_aCD3','1e_6h_IL12','1b_6h_aCD3')
                           
#Checks to see if row and column names match              
all(colnames(counts_matrix) %in% rownames(sample_info))
all(colnames(counts_matrix) == rownames(sample_info))

```{r DESeq, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#Starting to run DESeq - using Time_treatment

dds <- DESeqDataSetFromMatrix(countData = counts_matrix,
                              colData = sample_info,
                              design = ~ Time_treatment)

#Pre-filtering

smallestGroupSize <- 3
keep <- rowSums(counts(dds) >= 200) >= smallestGroupSize
ddskeep <- dds[keep,]

#Reference level

ddskeep$Time_treatment <- factor(dds$Time_treatment, levels = c("1h_PBS"))

#"6h_PBS", "1h_CD3", "6h_CD3", "1h_IL12", "6h_IL12", "1h_IL18", "6h_IL18"

#droplevel to check if samples are blank
ddskeep$Time_treatment <- droplevels(dds$Time_treatment)

ddskept <- DESeq(ddskeep)
ddskept
res <- results(ddskept)
res


resultsNames(ddskept)


#Order by pvalue 
resOrdered <- res[order(res$pvalue),]
resOrdered

#significance of 0.05
res05 <- results(ddskept, alpha=0.05)
summary(res05)
res05
res05 <- res05[order(res$pvalue),]

#MAplot for these results
plotMA(res05, ylim=c(-2,2))

plotMA(res05, ylim=c(-10,3))

idx <- identify(res05$baseMean, res05$log2FoldChange)
rownames(res05)[idx]

#8 = Ifng, 7 = Tnf



plotCounts(ddskept, gene=which.min(res$padj), intgroup="Time_treatment")

plotCounts(ddskept, "Ifng",intgroup="Time_treatment")
plotCounts(ddskept, "Tnf",intgroup="Time_treatment")

#A nicer plot
library("ggplot2")
d <- plotCounts(ddskept, "Ifng",intgroup="Time_treatment", 
                returnData=TRUE)
ggplot(d, aes(x=Time_treatment, y=count)) + 
  geom_point(position=position_jitter(w=0.1,h=0)) +
  scale_y_log10(breaks=c(0,1000,5000,10000,50000,100000,500000))

e <- plotCounts(ddskept, "Tnf",intgroup="Time_treatment", 
                returnData=TRUE)
ggplot(e, aes(x=Time_treatment, y=count)) + 
  geom_point(position=position_jitter(w=0.1,h=0)) +
  scale_y_log10(breaks=c(0,1000,5000,10000,50000))

```{r Comparisons between conditions, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#6h PBS vs 1h PBS
resPBS <- results(ddskept, alpha=0.05, contrast=c("Time_treatment","6h_PBS","1h_PBS"))
resPBS
summary(resPBS)
resPBS <- resPBS[order(res$pvalue),]

#adj p-value<0.05, LFC 1447 up, 1470 down

plotMA(resPBS, ylim=c(-3,7))

idx <- identify(resPBS$baseMean, resPBS$log2FoldChange)
rownames(resPBS)[idx]


#1h CD3 vs 1h PBS
res1hPBSCD3 <- results(ddskept, alpha=0.05, contrast=c("Time_treatment","1h_CD3","1h_PBS"))
res1hPBSCD3
summary(res1hPBSCD3)
res1hPBSCD3 <- res1hPBSCD3[order(res$pvalue),]

#adj p-value<0.05, LFC 55 up, 0 down

plotMA(res1hPBSCD3, ylim=c(-2,11))

idx <- identify(res1hPBSCD3$baseMean, res1hPBSCD3$log2FoldChange)
rownames(res1hPBSCD3)[idx]


#1h IL12 vs 1h PBS
res1hPBSIL12 <- results(ddskept, alpha=0.05, contrast=c("Time_treatment","1h_IL12","1h_PBS"))
res1hPBSIL12
summary(res1hPBSIL12)
res1hPBSIL12 <- res1hPBSIL12[order(res$pvalue),]

#adj p-value<0.05, LFC 26 up, 1 down

plotMA(res1hPBSIL12, ylim=c(-3,5))

idx <- identify(res1hPBSIL12$baseMean, res1hPBSIL12$log2FoldChange)
rownames(res1hPBSIL12)[idx]


#1h IL18 vs 1h PBS
res1hPBSIL18 <- results(ddskept, alpha=0.05, contrast=c("Time_treatment","1h_IL18","1h_PBS"))
res1hPBSIL18
summary(res1hPBSIL18)
res1hPBSIL18 <- res1hPBSIL18[order(res$pvalue),]

#adj p-value<0.05, LFC 0 up, 0 down

plotMA(res1hPBSIL18, ylim=c(-2,2))

idx <- identify(res1hPBSIL18$baseMean, res1hPBSIL18$log2FoldChange)
rownames(res1hPBSIL18)[idx]


#6h CD3 vs 6h PBS
res6hPBSCD3 <- results(ddskept, alpha=0.05, contrast=c("Time_treatment","6h_CD3","6h_PBS"))
res6hPBSCD3
summary(res6hPBSCD3)
res6hPBSCD3 <- res6hPBSCD3[order(res$pvalue),]

#adj p-value<0.05, LFC 3006 up, 2680 down

plotMA(res6hPBSCD3, ylim=c(-3,15))

idx <- identify(res6hPBSCD3$baseMean, res6hPBSCD3$log2FoldChange)
rownames(res6hPBSCD3)[idx]


#6h IL12 vs 6h PBS
res6hPBSIL12 <- results(ddskept, alpha=0.05, contrast=c("Time_treatment","6h_IL12","6h_PBS"))
res6hPBSIL12
summary(res6hPBSIL12)
res6hPBSIL12 <- res6hPBSIL12[order(res$pvalue),]

#adj p-value<0.05, LFC 684 up, 517 down

plotMA(res6hPBSIL12, ylim=c(-4,8))

idx <- identify(res6hPBSIL12$baseMean, res6hPBSIL12$log2FoldChange)
rownames(res6hPBSIL12)[idx]


#6h IL18 vs 6h PBS
res6hPBSIL18 <- results(ddskept, alpha=0.05, contrast=c("Time_treatment","6h_IL18","6h_PBS"))
res6hPBSIL18
summary(res6hPBSIL18)
res6hPBSIL18 <- res6hPBSIL18[order(res$pvalue),]

#adj p-value<0.05, LFC 4 up, 0 down

plotMA(res6hPBSIL18, ylim=c(-2,6))

idx <- identify(res6hPBSIL18$baseMean, res6hPBSIL18$log2FoldChange)
rownames(res6hPBSIL18)[idx]


#6h CD3 vs 1h CD3
res6hcd3CD3 <- results(ddskept, alpha=0.05, contrast=c("Time_treatment","6h_CD3","1h_CD3"))
res6hcd3CD3
summary(res6hcd3CD3)
res6hcd3CD3 <- res6hcd3CD3[order(res$pvalue),]

#adj p-value<0.05, LFC 3069 up, 3103 down

plotMA(res6hcd3CD3, ylim=c(-5,12))

idx <- identify(res6hcd3CD3$baseMean, res6hcd3CD3$log2FoldChange)
rownames(res6hcd3CD3)[idx]


#6h IL12 vs 1h IL12
res6hil12IL12 <- results(ddskept, alpha=0.05, contrast=c("Time_treatment","6h_IL12","1h_IL12"))
res6hil12IL12
summary(res6hil12IL12)
res6hil12IL12 <- res6hil12IL12[order(res$pvalue),]

#adj p-value<0.05, LFC 1881 up, 1792 down

plotMA(res6hil12IL12, ylim=c(-7,6))

idx <- identify(res6hil12IL12$baseMean, res6hil12IL12$log2FoldChange)
rownames(res6hil12IL12)[idx]


#6h IL18 vs 1h IL18
res6hil18IL18 <- results(ddskept, alpha=0.05, contrast=c("Time_treatment","6h_IL18","1h_IL18"))
res6hil18IL18
summary(res6hil18IL18)
res6hil18IL18 <- res6hil18IL18[order(res$pvalue),]

#adj p-value<0.05, LFC 1493 up, 1447 down

plotMA(res6hil18IL18, ylim=c(-7,5))

idx <- identify(res6hil18IL18$baseMean, res6hil18IL18$log2FoldChange)
rownames(res6hil18IL18)[idx]

  
```{r Heatmap DESeq, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

install.packages('pheatmap')
library("pheatmap")

ntd <- normTransform(ddskept)
#Need normalised data for the heatmap
#Can do these transformations too
vsd <- vst(ddskept, blind=FALSE)
rld <- rlog(ddskept, blind=FALSE)

select <- order(rowMeans(counts(ddskept,normalized=TRUE)),
                decreasing=TRUE)[1:20]
df <- as.data.frame(colData(ddskept)[,c("Time_treatment","Mouse")])
pheatmap(assay(ntd)[select,], cluster_rows=FALSE, show_rownames=TRUE,
         cluster_cols=FALSE, annotation_col=df)

#Can do these heatmaps with the other data transformations too
pheatmap(assay(vsd)[select,], cluster_rows=FALSE, show_rownames=TRUE,
         cluster_cols=FALSE, annotation_col=df)

pheatmap(assay(rld)[select,], cluster_rows=FALSE, show_rownames=TRUE,
         cluster_cols=FALSE, annotation_col=df)
         
#Sample to sample distances
sampleDists <- dist(t(assay(vsd)))

install.packages("RColorBrewer")
library("RColorBrewer")

sampleDistMatrix <- as.matrix(sampleDists)
rownames(sampleDistMatrix) <- paste(vsd$Sample_name, sep="-")
colnames(sampleDistMatrix) <- paste(vsd$Sample_name, sep="-")
colors <- colorRampPalette( rev(brewer.pal(9, "Blues")) )(255)
pheatmap(sampleDistMatrix,
         clustering_distance_rows=sampleDists,
         clustering_distance_cols=sampleDists,
         col=colors)

#PCA plot
plotPCA(vsd, intgroup=c("Time", "Treatment"))

plotPCA(vsd, intgroup=c("Time_treatment", "Replicate"))

plotPCA(vsd, intgroup=c("Sex", "Mouse"))

#Time and treatment
pcaData <- plotPCA(vsd, intgroup=c("Time", "Treatment"), returnData=TRUE)
percentVar <- round(100 * attr(pcaData, "percentVar"))
ggplot(pcaData, aes(PC1, PC2, color=Time, shape=Treatment)) +
  geom_point(size=3) +
  xlab(paste0("PC1: ",percentVar[1],"% variance")) +
  ylab(paste0("PC2: ",percentVar[2],"% variance")) + 
  coord_fixed()

#Sex and Mouse
pcaData2 <- plotPCA(vsd, intgroup=c("Sex", "Mouse"), returnData=TRUE)
percentVar2 <- round(100 * attr(pcaData2, "percentVar"))
ggplot(pcaData2, aes(PC1, PC2, color=Sex, shape=Mouse)) +
  geom_point(size=3) +
  xlab(paste0("PC1: ",percentVar2[1],"% variance")) +
  ylab(paste0("PC2: ",percentVar2[2],"% variance")) + 
  coord_fixed()

```{r Contrasts DESeq, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#All seems like a repeat of above
#In the MAplot section

results(ddskept, contrast=c("Time_treatment","1h_PBS" ,"1h_CD3" ))
#contrast = c(factor, numerator, denominator)

resultsNames(ddskept)
#"Intercept""Time_treatment_1h_IL12_vs_1h_CD3""Time_treatment_1h_IL18_vs_1h_CD3""Time_treatment_1h_PBS_vs_1h_CD3""Time_treatment_6h_CD3_vs_1h_CD3""Time_treatment_6h_IL12_vs_1h_CD3""Time_treatment_6h_IL18_vs_1h_CD3" "Time_treatment_6h_PBS_vs_1h_CD3" 

#Interactions

ddskept$Time_treatment <- factor(paste0(ddskept$Time, ddskept$Treatment))
design(ddskept) <- ~ Time_treatment
ddsK <- DESeq(ddskept)
resultsNames(ddsK)

#Potentially have to do all "1hPBS", "6hPBS", "1hCD3", "6hCD3", "1hIL12", "6hIL12", "1hIL18", "6hIL18"
results(ddsK, contrast=c("Time_treatment", "1hPBS", "1hCD3"))

PCa <- plotCounts(ddsK, 'Ifng', intgroup="Time_treatment", 
                returnData=TRUE)
                
ggplot(PCa, aes(x=Time_treatment, y=count)) + 
  geom_point(position=position_jitter(w=0.1,h=0)) + scale_y_log10(breaks=c(0, 1000,5000,10000,50000,100000))



```{r multifactor DESeq, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
#Don't think this works for the data set
#ends up being the same as the analysis above

colData(dds)
ddsMF <- dds

levels(ddsMF$Time_treatment)

levels(ddsMF$Time_treatment) <- sub("_-.*", "", levels(ddsMF$Time_treatment))
levels(ddsMF$Time_treatment)

design(ddsMF) <- formula(~ Time_treatment)
ddsMF <- DESeq(ddsMF)

resMF <- results(ddsMF)
head(resMF)

```{r Other DESeq, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
#Need to run this again to check it all and compare it to the analysis above

dds <- DESeqDataSetFromMatrix(countData = counts_matrix,
                              colData = sample_info,
                              design = ~ Time_treatment)

ddsDE <- DESeq(dds)

resultsNames(ddsDE)

keepdds_200 <- rowSums(counts(dds)) >= 200

ddskeep <- dds[keepdds_200,]

dds$Time_treatment <- factor(dds$Time_treatment, levels = c("PBS"))

ddsDET_T <- DESeq(dds2)
resT_T <- results(ddsDET_T)

summary(resT_T)

resT_T0.05 <- results(ddsDET_T, alpha = 0.05)
summary(resT_T0.05)


```{r DESeq with Time and Treatment, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
#Need to run this again to check it all and compare it to the analysis above

dds2 <- DESeqDataSetFromMatrix(countData = counts_matrix,
                              colData = sample_info,
                              design = ~ Time + Treatment)

#Pre-filtering

smallestGroupSize <- 4
keep2 <- rowSums(counts(dds2) >= 200) >= smallestGroupSize
dds2keep <- dds2[keep2,]

#Reference level

resultsNames(dds2kept)

dds2keep$Time <- factor(dds2$Time, levels = c("1h"))
#"6h_PBS", "1h_CD3", "6h_CD3", "1h_IL12", "6h_IL12", "1h_IL18", "6h_IL18"

#droplevel to check if samples are blank
dds2keep$Time_treatment <- droplevels(dds2$Time_treatment)

dds2kept <- DESeq(dds2keep)
dds2kept
res2 <- results(dds2kept)
res2

resultsNames(dds2kept)

#significance of 0.05
res205 <- results(dds2kept, alpha=0.05)
summary(res205)
res205
res205 <- res205[order(res$pvalue),]


#MAplot for these results
plotMA(res205, ylim=c(-2,2))

plotMA(res205, ylim=c(-15,2))

idx <- identify(res205$baseMean, res205$log2FoldChange)
rownames(res205)[idx]

vsd2 <- vst(dds2kept, blind=FALSE)


plotPCA(vsd2, intgroup=c("Time", "Treatment"))

plotPCA(vsd2, intgroup=c("Sex", "Mouse"))




ddsDE2 <- DESeq(dds2)
resultsNames(ddsDE2)

keepdds2_200 <- rowSums(counts(dds2)) >= 200
dds2keep <- dds2[keepdds2_200,]

dds2$Time_Treatment <- factor(dds$Time_treatment, levels = c("PBS"))

#Filtered to remove any genes with fewer than 200 counts

dds2keep$Time_Treatment <- factor(dds$Time_treatment, levels = c("PBS"))
dds2keepDET_T <- DESeq(dds2keep)
res2keepT_T0.05 <- results(dds2keepDET_T, alpha = 0.05)
summary(res2keepT_T0.05)

```{r cars}

```

## Including Plots

You can also embed plots, for example:

```{r pressure, echo=FALSE}

```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
